Relationships in the extremes and their influence on competition and coexistence

### 2021-12-30 paper/Paper.Rnw

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

%***DAN: Methods paras here for G4, if we accomplish G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species coexistence which highlights the role ATAs can play, we consider the coexistence of diatom species in chemostat batch experiments by Descamps-Julien & Ganzalez 2005. In their experiments, measurements of growth of Cyclotella pseudostelligera and Fragilaria crotonensis over a range of temperatures was used to parametrize a competitive resource model as follows:
put model here
Descamps-Julien & Gonzalez used simulations to show that coexistence occurs with oscillating temperature fluctuations with parameters $\theta_0 = 18, a = 6, P = 60$d) and Ellner 2016 and 2018 broke down their coexistence into the contrirbuting mechanisms. Here, we consider varying parameters of temperature fluctuations and further decompose the coexistence of C. pseudostelligera and F. crotonensis to account for contributions of ATAs and ATA effects as described above.

Like for G2, we first plotted the terms of equation \ref{eq:full_decomp} for the chemostat model with $E=$ and $C=$ against independent parameters of temperature oscillation, $\theta_0, a$, and $P$, a total of three plots. For each plot, we set two of the parameters constant at the original parameters which coexistence was found in Descamps-Julien & Gonzalez, and let the remaining temperature parameter vary. %say which scenario this is in Ellner 2018
We let $\theta_0$ vary between $[x,y]$, $a$ vary between $[x,y]$, and $P$ vary between $[x,y]$. 
Relative importance of contributions was judged from the 
plots to assess whether and to what extent ATA effects on coexistence were meaningfully large for the lottery model, compared to 
the contributions of other mechanisms. 
The simulation methods of \cite{Ellner_2019} were used to compute the components of (\ref{eq:full_decomp}).

To further explore the parameter space, we generated 2 dimensional surface plots which reveal the subsets of parameter space where coexistence is possible with ranging degrees of ATA contributions and where ATA rescue occurs. We evaluated coexistence under two varying parameters and indicated $\Delta_i^{[EC]}/IGR$, a ratio of ATA contribution to total IGR, with a color bar and contour lines. Positive values of $\Delta_i^{[EC]}/IGR$ indicate coexistence and negative values indicate failure of coexistence. Greater magnitudes of $\Delta_i^{[EC]}/IGR$ and higher color intensity indicate greater contribution of ATAs to the IGR and positive cases where $\Delta_i^{[EC]}>IGR$ indicate ATA rescue.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{The contribution of ATAs to diatom coexistence is meaningfully large compared to 
other components of the MCT decomposition. All panels show the IGR and the decomposition of
the IGR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel, 
two of the three parameters defining temperature oscillation, amplitude, period, and mean, 
is fixed at the original coexistence values found by Descamps-Julien & Gonzalez ($a=6, P=60, Tbar=18$). The IGR and components of the decomposition
are plotted against one varying parameter. The original value of the varying parameter and coexistence quantities at that value is indicated by the vertical dashed magenta lines. The contribution of ATAs (red line)
has magnitude comparable to other components and, in these parameters, is always positive, indicating that ATAs can contribute meaningfully
to diatom coexistence. In some instances, the IGR line (orange) falls 
between the ATA line (red)
and the $x$ axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of the weaker competitor. Standard errors of plotted quantities ?.  \label{fig:fig5}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig6.pdf}
\caption{The degree of ATA contribution to the IGR change with parameters of temperature oscillation. The ratio $\Delta_i^{[EC]}/IGR$ constitutes the degree of ATA contribution to thee IGR with larger magnitudes signifying larger ATA contributions. In this model, for the plotted parameters, positive $\Delta_i^{[EC]}/IGR$ only result from positive $\Delta_i^{[EC]}$ and positive $IGR$ and negative $\Delta_i^{[EC]}/IGR$ only result from positive $\Delta_i^{[EC]}$ and negative $IGR$. Thus, cyan areas indicate failure to coexist and magenta areas indicate coexistence with ATA contribution increasing with color intensity. The most intense areas of magenta indicate parameter subspaces where ATA rescue occurs. For simplicity and readability magnitudes of $\Delta_i^{[EC]}/IGR$ >1 are capped at $\abs{1}$. } \label{fig:fig6}
\end{figure}

>>>>

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Descamps_2005, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

%***DAN: Methods paras here for G4, if we accomplish G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species coexistence which highlights the role ATAs can play, we consider the coexistence of diatom species in chemostat batch experiments by \cite{Descamps_2005}. In their experiments, measurements of growth of \emph{Cyclotella pseudostelligera} and \emph{Fragilaria crotonensis} over a range of temperatures was used to parametrize a competitive resource model as follows:
put model here
\cite{Descamps_2005} used simulations to show that coexistence occurs with oscillating temperature fluctuations with parameters $\theta_0 = 18, a = 6, P = 60$d) and \cite{Ellner_2016, Ellner_2019} broke down their coexistence into the contrirbuting mechanisms. Here, we consider varying parameters of temperature fluctuations and further decompose the coexistence of \emph{C. pseudostelligera} and \emph{F. crotonensis} to account for contributions of ATAs and ATA effects as described above, using decomposition defined by $E=...$ and $C=...$.

Like for G2, we first plotted the terms of equation \ref{eq:full_decomp} for the chemostat model with $E=$ and $C=$ against independent parameters of temperature oscillation, $\theta_0, a$, and $P$, a total of three plots. For each plot, we set two of the parameters constant at the original parameters which coexistence was found in \cite{Descamps_2005}, and let the remaining temperature parameter vary. %say which scenario this is in Ellner 2018
We let $\theta_0$ vary between $[x,y]$, $a$ vary between $[x,y]$, and $P$ vary between $[x,y]$. 
Relative importance of contributions was judged from the 
plots to assess whether and to what extent ATA effects on coexistence were meaningfully large for the lottery model, compared to 
the contributions of other mechanisms. 
The simulation methods of \cite{Ellner_2019} were used to compute the components of (\ref{eq:full_decomp}).

To further explore the parameter space, we generated 2 dimensional surface plots which reveal the subsets of parameter space where coexistence is possible with ranging degrees of ATA contributions and where ATA rescue occurs. We evaluated coexistence under two varying parameters and indicated $\Delta_i^{[EC]}/\text{IGR}$, a ratio of ATA contribution to total IGR, with a color bar and contour lines. Positive values of $\Delta_i^{[EC]}/\text{IGR}$ indicate coexistence and negative values indicate failure of coexistence. Greater magnitudes of $\Delta_i^{[EC]}/IGR$ and higher color intensity indicate greater contribution of ATAs to the IGR and positive cases where $\Delta_i^{[EC]}>IGR$ indicate ATA rescue.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{The contribution of ATAs to diatom coexistence is meaningfully large compared to 
other components of the MCT decomposition. All panels show the IGR and the decomposition of
the IGR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel, 
two of the three parameters defining temperature oscillation, amplitude, period, and mean, 
is fixed at the original coexistence values found by \cite{Descamps_2005} ($a=6, P=60, \theta_0=18$). The IGR and components of the decomposition
are plotted against one varying parameter. The original value of the varying parameter and coexistence quantities at that value is indicated by the vertical dashed magenta lines. The contribution of ATAs (red line)
has magnitude comparable to other components and, in these parameters, is always positive, indicating that ATAs can contribute meaningfully
to diatom coexistence. In some instances, the IGR line (orange) falls 
between the ATA line (red)
and the $x$ axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of the weaker competitor. Standard errors of plotted quantities ?.}  \label{fig:fig5}
\end{figure}

\begin{figure}
\includegraphics[width=.38\textwidth]{../results_figs/fig6.pdf}
\caption{The degree of ATA contribution to the IGR change with parameters of temperature oscillation. The ratio $\Delta_i^{[EC]}/\text{IGR}$ constitutes the degree of ATA contribution to thee IGR with larger magnitudes signifying larger ATA contributions. In this model, for the plotted parameters, positive $\Delta_i^{[EC]}/\text{IGR}$ only result from positive $\Delta_i^{[EC]}$ and positive IGR and negative $\Delta_i^{[EC]}/\text{IGR}$ only result from positive $\Delta_i^{[EC]}$ and negative IGR. Thus, cyan areas indicate failure to coexist and magenta areas indicate coexistence with ATA contribution increasing with color intensity. The most intense areas of magenta indicate parameter subspaces where ATA rescue occurs. For simplicity and readability magnitudes of $\Delta_i^{[EC]}/\text{IGR}$ >1 were capped at $|1|$. } \label{fig:fig6}
\end{figure}
